Writing Trash Bin

### 2024-08-05 paper.qmd
wrote value extraction bit

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
Therefore the best uncertainty visualisation is not am "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data, that gives us a good idea of the limitations of our model.

@@

# Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", are lacking if we want any generalizable results. However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016]. Additionally, several authors have linked this problem to a  "task" or "goal" dependence in the evaluations.

Here we discuss the cause of this goal dependence, the current methods for evaluating uncertainty visualisation, and we ask if these methods are sufficient to evaluate graphics on their ability to perform singal-suppression.

## The Hermann grid illusion of visualisation experiments
When you look at a graphic, what statistics are you inferring from the visulisation? Depending on the graphic, you might extract several statistics at once. However, this is assuming I have just presented a graphic and said nothing. What is I asked you to extract the mean of $X$? Then it might be safe to assume you are doing inference on $\bar{X}$.

By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].

Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.

A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them. This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.

Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.  @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.

As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise. Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.

## Current methods

@Kinkeldey2014 noted that most studies assessed users ability to extract the estimate and the uncertainty as separate signals, and expressed doubt that this was an effective evaluation tool for conveying their "single integrated value".

There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors. These secondary benefits are, however, *not* the immediate goal of uncertainty. The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.

### Value extraction

It is unclear if this sufficient to assess a visualisation on its ability to communicate a

There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).

Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].

Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat. For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. @Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.

The final method used by authors is to just explicitly ask about uncertainty and signal information separately. @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty. Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables. @Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations. These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together. Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.

@@

### Vague questions

@@

## Testing signal supression
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

>>>>

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.

This relationship between uncertainty and the "purpose" of our analysis is littered throughout the literature. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.  @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.

This makes it very difficult to move quantified uncertainty through the the layers of our analysis, especially when designing a visualisation for EDA. If we don't know what the final statistic is, we cannot quantify the effects of our assumptions. Therefore, often the best uncertainty visualisation is not an "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data as it gives us a good idea of it's limitations.

@@

# Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", are lacking if we want any generalizable results. However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].

Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consier how we to best evaluate visualisations on their ability to perform signal-suppression.

## Current methods
Including uncertainty in a visualisation comes with many secondary benefits. Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance. Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.

### Value extraction of uncertainty statistics
Uncertainty visualisations are most commonly evaluated on how accurately readers can extract information from them [@Hullman2019]. This most commonly comes in the form of separate questions about estimates variance [@Kinkeldey2014]. This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". Additionally, uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] regardless of what information can actually be extracted by the chart. This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? Uncertainty is not something we are trying to infer, uncertainty is a by-product of inference. By shifting the focus of our inference to $Var(X)$ or $P(X)$ we are just evaluating visualisations on their ability to convey uncertainty statistics. These graphics may not always set out to communicate uncertainty as a signal, but they are certainty evaluating uncertainty as a signal.

The problem with evaluating uncertainty as a signal are identical to the problems associated with displaying uncertainty as a signal. There is no reason to assume uncertainty would behave any differently to any other variable when we evaluate them in this way. For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. @Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. It is well established that extracting information from a graphic using a perceptual task will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].

Directly asking about uncertainty evaluates uncertainty as a signal, not as noise, so it is not a reliable evaluation method when we are designing graphics for signal suppression. Many uncertainty visualisation researchers are already aware of this, so direct questions are not the only methods used to evaluate uncertainty visualisations. We will discuss these alternatives in the following section.

@@

### Vague questions
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@@

## Testing signal supression

If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
- trust experiment
- weird animation of grass

Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion. A simple example is the spatial trend in

If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

### 2024-08-05 writingdumpster.qmd
wrote value extraction bit

# Removed Sections
@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".

@@

>>>>

# Removed Sections
The beauty of visualisation is that it *tells* us what we should be concerned about in our analysis. This aspect disappears the second we ask a direct question about our visualisation.

@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".

@@

# Removed comments on experiments
The final method used by authors is to just explicitly ask about uncertainty and signal information separately. @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty. Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables. @Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations. These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together. Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.

# Alternative purpose section
This purpose or task dependency appears repeatedly in the literature, although its connection to inference is seldom mentioned. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. This is because "signal" or "noise" are not inherent qualities of our data, they are simply the terms we give to the information we wants to keep, and the information we want to throw away. Whether or not information is signal or noise depends entirely on the questions we are are seeking to answer [@Meng2014]. This distinction between noise and signal occurs as early as data collection, where the entire process of deciding how to boil real world entities down into data depends on the relationships we seek to identify [@Otsuka2023]. This is true for our entire analysis, right up to deciding how to display information, as the way we display information should be based on that the user is supposed to do with it [@Fischhoff2014]. Even the process we use to combine uncertainty from different stages of an analysis depends on why the uncertainty matters [@Wallsten1997]. It is clear that as we move through the stages of our analysis noise is the status given to the information we throw away that inhibits our ability to identify signal. We want to reintroduce some noise back into our visualisation such that we can perform signal-suppression.

# Example for perceptual task problem
Consider the typical uncertainty visualisations shown in @fig-mean, which includes a density plot, a sample, and an mean with confidence intervals. These graphics each show the distribution of the standard normal random variable X using a density plot, a sample, and a mean with confidence bars. Each of these technically show a piece of information with zero uncertainty. The density plot provides a deterministic interpretation of the PDF, the sample shows exactly the values we observed, and the mean with error bars shows three explicitly calculated statistics ($X_{0.05}, \bar{x}, X_{0.95}$). If we were to use these visualisations to extract the mean, the role of signal in the chart shifts, and instead of directly reading the values, we need to do mental calculations. The density plot requires us to find the midpoint along the x axis that could spit the area under the density curve perfect in half. The sample requires us to find the point that minimizes the sum of squares errors, i.e. the total lengths from each point to the mean. The mean with error bars allows us to simply read the mean off the graphic using a basic perceptual task. If we were to use the visualisations to extract the variance instead, we could describe a similar process.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-mean
#| fig-cap: "This visualisation "
#| fig-subcap: 
#|   - "Probability density function"
#|   - "Sample"
#|   - "Mean with error bars"
#| layout-ncol: 3
#| layout-valign: "bottom"
set.seed(1)
xdata <- tibble(x=rnorm(100))

# Density function
x1 <- xdata |>
  ggplot(aes(x=x)) +
  geom_density(colour="#1e81b0", fill="#abdbe3", alpha=0.5) + 
  scale_x_continuous(name = "x") +
  scale_y_continuous(name = "P(X)") +
  theme_minimal() + 
  theme(aspect.ratio=1)

# Sample

# Mean with error bars
x3 <- xdata %>%
  summarise(avg = mean(x),
          conf_95a = quantile(x, probs=c(0.025)),
          conf_95b = quantile(x, probs=c(0.975))) %>%
  ggplot(aes(y="NA")) +
  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1, colour="#abdbe3") +
  geom_point(aes(x=avg), colour="#1e81b0") +
  scale_x_continuous(name = "X",
                     breaks = seq(-2,2, 0.5)) +
  theme_minimal() +
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=1)

```
